Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents

### 2025-07-24 doc/cicm25-paper/paper.tex
some more paper editing

\title{
    % Semantic Authoring in the \sTeX/\ALeA World
    % \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''}
        Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents
}
\titlerunning{
    Semantic Authoring in a Flexiformal Context \textellipsis
}

@@

% \keywords{semantic authoring, \snify, \sTeX, \ALeA, flexiformal content commons}
\institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
\begin{document}
\maketitle
\begin{abstract}
%   Semantic annotations can be utilized for learning support services in technical
%   documents. The resulting active documents support more direct and personalized
%   interaction with the knowledge conveyed, but require a large investment in the authoring
%   process.
% 
%   In this paper we discuss the nature of the semantic authoring problem and argue that
%   while sharing aspects of classical, informal authoring and formal authoring, it is a
%   problem of its own that needs special editing support facilities. We present a simple
%   utility that supports the process of annotating technical terms with semantic references
%   to a flexiformal content commons that enhances annotation productivity by almost an
%   order of magnitude.

  Semantic annotations enable services that make the knowledge in documents more
  accessible.  In the case of \ALeA, an adaptive learning assistant we use as a case study, they
  enable, for example, guided tours and practice problems tailored to the learner's
  competencies.  However, the cost of semantic authoring is substantial and tool support
  is limited.

  In this paper we discuss the nature of the semantic authoring problem and argue that,
  while sharing aspects of both classical (informal) authoring and formal authoring, it is
  a problem of its own that needs special editing support facilities.  We present and
  evaluate a simple utility that supports the process of annotating technical terms with
  semantic references to a flexiformal content commons and increases annotation
  productivity by almost an order of magnitude.
\end{abstract}
% \setcounter{tocdepth}{3}
% \tableofcontents

@@

As a case study, we will use adaptive learning assistants for STEM subjects in higher
education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.  We contend that
machine-learning based technologies are largely unsuitable for this task -- at least until
they reach the precision and explainability required by the ethics of
teaching\edited{\footnote{Incidentally, the AI Act of the European Union \cite{EUAIAct:on}
    classifies education as a high-risk area and thus mandates high requirements of
    explainability, accountability, and human oversight for educational AI applications.}}
-- while the scope of specific domains is just narrow enough that symbolic approaches are
viable.  Conventional (\LaTeX-based) lecture materials can essentially be directly
imported into \ALeA, but its ability to generate learner-adaptive learning objects --
instrumented with learning support interactions -- depends on semantic annotations.  These
can be embedded in \LaTeX\ documents using the \sTeX
package~\cite{MueKo:sdstex22,sTeX:github:on}.  In this paper, we will mostly focus on
annotating terms with the semantic concepts they refer to, but \sTeX supports a much wider
range of annotations, such as annotating problems with the learning \edited{prerequisites
  and} objectives or annotating the defining phrase (definiens) in concept definitions,
and could, in principle, even be used for full formalization.  As \sTeX supports the
spectrum from informal (un-annotated) to fully formalized content, we say that \sTeX
content is \textbf{flexiformal}.

@@

\paragraph{Overview and Contribution}
In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\Cref{sec:semauth}).
Using \sTeX/\ALeA as an exemplary case study,
we will provide an overview of the authoring support (\Cref{sec:tools})
and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\Cref{sec:snify}).
Afterwards, we discuss on-going work to integrate \snify better
into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
The paper ends with a brief discussion of related work (\Cref{sec:relwork}) and the conclusion (\Cref{sec:conclusion}).

@@

\begin{figure}[h]\centering
  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
\end{figure}

@@

To annotate the word ``terminal'' in the document, the author has to 
\begin{inparaenum}[\em i\rm)]
\item be aware that it is a
technical term, 
\item that the module from \Cref{fig:state-space} exists,
\item know the symbol URI, 
\item manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
  directive if it is not already (recursively) implied and possibly removing directives
  that become redundant by the new one.
\end{inparaenum}
\edited{A rather daunting task for a lecture with 50,000 -- 100,000 words in notes and slides and ca. 10--15\,\% of them technical terms.\footnote{
    \edited{
        Estimates based on some of our own lectures materials
        ranging from a small introductory lecture (ca. 30,000 words)
        to a large two-semester lecture on AI (ca. 140,000 words).
        The 10--15\,\% estimate appears to be relatively consistent across our lectures
        but depends on the author's intuition about what a technical term is.
        Other lecturers or lectures on other topics may have very different values.
    }
}}

@@

\edited{The annotation catalog is language-specific.}
% For documents in a language $L$,
% where $L$ is not English, the annotation practices are a bit
% more complex still, in particular because the annotation catalog is usually incomplete \edited{for $L$}.
% Note that for $L$-annotation, we need an $L$-specific annotation catalog. 
\sTeX already supports multilinguality:
\begin{itemize}
\item Documents have explicit language annotations (usually in form of the file extension
  \lstinline[mathescape]|.$\llangle lang\rrangle$.tex| where $\llangle lang\rrangle$ is the
  ISO-639 language identifier).
\item \edited{Modules in the domain model allow translations that share the formal
    declarations with another module.}
        % In particular all definienda in $L$-modules are $L$-specific.
\end{itemize}
% But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
\edited{In practice, the annotation catalog for languages other than English is often inadequate as}
\begin{enumerate}[\em i\rm)]
    \item \edited{the domain model may not have translations in that language -- especially for the ``smaller'' languages, and}
    \item \edited{symbol names, usually English words, cannot be used as a verbalizations.}
\end{enumerate}
\edited{In this situation, we usually have to invest in
establishing a suitable catalog -- e.g. by providing translations (or at least dictionary
information) in the domain model or
delineating modules and annotating definienda that the
domain model does not cover in the documents to be annotated.
}

\subsection{Semantic Authoring as a Distinct Problem}
We contend that semantic authoring is a problem that is distinct from traditional
(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
while sharing some of the aspects of either.

\paragraph{Traditional, informal authoring}
In traditional authoring, e.g. of a scientific article or a textbook, the domain model --
a highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal (usually a natural language like English
augmented with technical jargon, formulae, tables, and diagrams) and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors. Consequently a large part of the
authoring problem lies in predicting the state and extent of the domain model of the
reader and creating text that bridges the gap between the reader's domain model and the
payload content of the article -- the knowledge the article or textbook intends to impart
-- ideally using the technical vocabulary the reader is familiar with to reduce this
gap. Tool support for traditional authoring is minimal and usually restricted to
spell/grammar-checkers and online thesauri if we disregard LLM-based writing/formulation
support for now.

@@

In addition to the base text, semantic authoring also has to provide the semantic
annotations that drive the personalization and interaction. Techniques from formal
authoring could help here, but in contrast to formal authoring, we cannot just trigger the
IDE functionality by the formal syntax, as the authored content is -- so far -- informal
and the content commons (the target of the annotations) is itself only flexiformal. In a
way, we need a local flexiformalization workflow that helps authors bridge the
informal-formal gap.

A distinctive feature of semantic authoring that is not apparent at the first glance is
that the added investment in annotation moves the sweet spot in the re-use/customize
trade-off towards sharing and interoperability. In our experience sufficiently annotated
documents can be automatically customized if the semantics has multiple
notations/verbalizations/presentations. Consider for instance course materials on complex
arithmetic, which become re-usable across the electrical engineering/mathematics border if
the imaginary unit in formulae is represented e.g. by a semantic macro
\lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths) and
$j$ (for EE uses). As a consequence semantic authoring is more geared towards creating
reusable/shareable document fragments organized in central or federated repositories and
hubs. This is quite similar to the formal authoring arena, where we see a tendency towards
massively shared math libraries in formal systems and package/library hubs for programming
libraries. Semantic authoring support systems should therefore facilitate the
logistics of using and curating shared libraries.

@@

How the annotation process can be supported depends on the particular semantic framework,
the specific annotations we want to support, and the semantics embedded in the content
commons.  In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers
to their definienda.  A tool for supporting semantic authoring of formulae via notations
has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.  It uses a
grammar to parse informal, presentation-focused formulae (e.g.\
\lstinline[keywordstyle={}]|A\times B|) into abstract syntax trees based on semantic
macros defined in \sTeX, which can then replace the original formula (e.g.\
\lstinline|\cart{A,B}| for the Cartesian product of sets).  In the case of ambiguity,
authors can pick the correct reading in a graphical user interface that visualizes the
different abstract syntax trees.

@@

To skip the word, the user can type \lstinline|s|;
\lstinline|s!| skips it until the end of the file, and \lstinline|S|
will skip it in future runs as well by appending a comment to the file.
There are numerous other commands, e.g.\ to adjust the selection,
search for alternative annotation targets,
or view/edit the file introducing one of the choices.

@@

\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents.}
As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
documents that were originally authored without semantic annotations.  It can be
configured to consider different \LaTeX\ macros to identify definienda (e.g.\
\lstinline[language={}]|\emph| or \lstinline[language={}]|\textbf|), and then steps
through them as \snify does.  For each potential definiendum, the user can then use fuzzy
search to find the corresponding symbol in the domain model.  This populates the catalog
with verbalizations relevant to the document, making subsequent \snify runs more
effective, especially for documents in languages that are not well covered by the domain
model (i.e.\ not English).

\paragraph{Evaluation}
% \section{Practical Evaluation and Future Work}\label{sec:eval}
\begin{wrapfigure}{r}{0.4\textwidth}
    \vskip-1.2cm
  \centering
  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
    \vskip-0.2cm
  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
  \vskip-0.7cm
\end{wrapfigure}
In our experience, \snify's step-through workflow for annotating term references 
% for symbols from the domain model 
is almost an order of magnitude more efficient than writing the
annotations and imports by hand.\footnote{
    \edited{
        The precise productivity gain depends on the annotator's
        familiarity with the domain model (for the right symbol
        is time-consuming) but also on other factors like
        the typing speed.
        Measurements are complicated by the fact that
        authors tend to annotate more terms when using \snify.
        We have observed speed-ups over IDE-based annotation
        ranging from 3 (for an expert knowing the domain model by heart)
        to 13 (for a beginner).
        The focus mode further increases productivity.
    }
}
% , even for annotators who are familiar with the domain model.
Given the sheer number of term annotations (see \Cref{fig:annocounts}),
this results in a considerable productivity gain.
For annotators unfamiliar with the domain model, the unassisted annotation task is
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices
% \ednote{
%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
% }
-- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

For less complete domain models we have to skip too many terms that should ultimately be
annotated and annotation effectiveness suffers. This is currently the case for all
non-English languages in the \sTeX corpus we work with. Coverage in German is only about
half of that of English and we can already see the practical effects.  The \defianno tool
can mitigate this partially by providing verbalizations for
the definienda in the document, but it does not help with terms that are not explicitly
defined in the document.

@@

\snify's interface is inspired by traditional spell checkers
like \lstinline|ispell|~\cite{ispellman},
which step through documents and suggest corrections for misspelled words.

The bulk annotation workflow for existing informal text is also relevant when creating
machine learning datasets, e.g.\ for named entity recognition.  There are many (often
browser-based) tools for this task; a prominent example is
\textsf{brat}~\cite{brat:on}.

In all cases, the setting, solution, and resulting workflows differ from the one presented
in this paper. The WissKI system \cite{goerz2010adaptation} has a rudimentary text edition
facility that allows annotating documents with ontology -- mostly ABox -- references; this
is the nearest match to our \snify system.

\section{Conclusion}\label{sec:conclusion}
In this paper we have presented the semantic authoring problem in a flexiformal context
using the \sTeX corpus and the adaptive learning assistant \ALeA as a concrete example. We
have shown that it differs from both classical authoring and formal authoring in terms of
the necessary system support. As an example of specialized authoring support, we have
presented \snify, a command-line tool that bridges the gap between informal text and
formal symbols in \sTeX.  Given the impact of \snify on the authoring process, we are
currently working on a deeper integration of \snify into the \sTeX ecosystem and content
commons as described in \Cref{sec:scaling}.

The \snify code is open source; the source and documentation are available from
\cite{stextools:git}.

>>>>

\title{
        Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents
}
\titlerunning{
    Semantic Authoring in a Flexiformal Context \textellipsis
}

@@

\institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
\begin{document}
\maketitle
\begin{abstract}
  Semantic annotations enable services that make the knowledge in documents more
  accessible.  In the case of \ALeA, an adaptive learning assistant we use as a case study, they
  enable, for example, guided tours and practice problems tailored to the learner's
  competencies.  However, the cost of semantic authoring is substantial and tool support
  is limited.

  In this paper we discuss the nature of the semantic authoring task and argue that,
  while sharing aspects of both classical (informal) authoring and formal authoring, it is
  a task of its own that needs special editing support facilities.  We present and
  evaluate a simple utility that supports the process of annotating technical terms with
  semantic references to a flexiformal content commons and increases annotation
  productivity by almost an order of magnitude.
\end{abstract}

\keywords{\edited{semantic authoring} \and flexiformality \and annotation \and \snify \and \sTeX \and \ALeA}
% \setcounter{tocdepth}{3}
% \tableofcontents

@@

As a case study, we will use adaptive learning assistants for STEM subjects in higher
education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.  We contend that
machine-learning based technologies are largely unsuitable for this task -- at least until
they reach the precision and explainability required by the ethics of
teaching\footnote{Incidentally, the AI Act of the European Union \cite{EUAIAct:on}
    classifies education as a high-risk area and thus mandates high requirements of
    explainability, accountability, and human oversight for educational AI applications.}
-- while the scope of specific domains is just narrow enough that symbolic approaches are
viable.  Conventional (\LaTeX-based) lecture materials can essentially be directly
imported into \ALeA, but its ability to generate learner-adaptive learning objects --
instrumented with learning support interactions -- depends on semantic annotations.  These
can be embedded in \LaTeX\ documents using the \sTeX
package~\cite{MueKo:sdstex22,sTeX:github:on}.  In this paper, we will mostly focus on
annotating terms with the semantic concepts they refer to, but \sTeX supports a much wider
range of annotations, such as annotating problems with the learning prerequisites
and objectives or annotating the defining phrase (definiens) in concept definitions,
and could, in principle, even be used for full formalization.  As \sTeX supports the
spectrum from informal (un-annotated) to fully formalized content, we say that \sTeX
content is \textbf{flexiformal}.

@@

\paragraph{Overview and Contribution}
In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct task requiring special tool support (\Cref{sec:semauth}).
Using \sTeX/\ALeA as an exemplary case study,
we will provide an overview of the authoring support (\Cref{sec:tools})
and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\Cref{sec:snify}).
Afterwards, we discuss on-going work to integrate \snify better
into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
The paper ends with a brief discussion of related work (\Cref{sec:relwork}) and the conclusion (\Cref{sec:conclusion}).

@@

\begin{figure}[h]\centering
  \fbox{\includegraphics[width=10cm,trim=0.8cm 0 0 0]{../img/minimax-remarks.pdf}}
  \caption{An unannotated learning object from an AI lecture}\label{fig:lo}
\end{figure}

@@

To annotate the word ``terminal'' in the document above, the author has to 
\begin{inparaenum}[\em i\rm)]
\item be aware that it is a
technical term, 
\item that the module from \Cref{fig:state-space} exists,
\item know the symbol URI, 
\item manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
  directive if it is not already (recursively) implied and possibly removing directives
  that become redundant by the new one.
\end{inparaenum}
A rather daunting task for a lecture with 50,000 -- 100,000 words in notes and slides and ca. 10--15\,\% of them technical terms.\footnote{
        Estimates based on some of our own lectures materials
        ranging from a small introductory lecture (ca. 30,000 words)
        to a large two-semester lecture on AI (ca. 140,000 words).
        The 10--15\,\% estimate appears to be relatively consistent across our lectures
        but depends on the author's intuition about what a technical term is.
        Other lecturers or lectures on other topics may have very different values.
}

@@

The annotation catalog is naturally language-specific.
\sTeX already supports multilinguality:
\begin{itemize}
\item Documents have explicit language annotations (usually in form of the file extension
  \lstinline[mathescape]|.$\llangle lang\rrangle$.tex| where $\llangle lang\rrangle$ is the
  ISO-639 language identifier).
\item Modules in the domain model include translations that share the formal
    declarations with another module.
        This induces a ``translation relation'' for verbalizations
        in different languages with the same symbol URI.
\end{itemize}
In practice, the annotation catalog for languages other than English is often inadequate as
\begin{enumerate}[\em i\rm)]
    \item the domain model may not have translations in that language -- especially for the ``smaller'' languages, and
    \item symbol names, usually English words, cannot be used as a verbalizations.
\end{enumerate}
In this situation, we usually have to invest in
establishing a suitable catalog -- e.g. by providing translations (or at least dictionary
information) in the domain model or
delineating modules and annotating definienda that the
domain model does not cover in the documents to be annotated.

\subsection{Semantic Authoring as a Distinct Task}
We contend that semantic authoring is a task that is distinct from traditional
(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
while sharing some of the aspects of either.

\paragraph{Traditional, informal authoring}
In traditional authoring, e.g. of a scientific article or a textbook, the domain model --
a highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal (usually a natural language like English
augmented with technical jargon, formulae, tables, and diagrams) and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors. Consequently a large part of the
authoring challenge lies in predicting the state and extent of the domain model of the
reader and creating text that bridges the gap between the reader's domain model and the
payload content of the article -- the knowledge the article or textbook intends to impart
-- ideally using the technical vocabulary the reader is familiar with to reduce this
gap. Tool support for traditional authoring is minimal and usually restricted to
spell/grammar-checkers and online thesauri if we disregard LLM-based writing/formulation
support for now.
\edited{
In our experience, LLM-based authoring support is less helpful in technical documents
than in everyday writing.
}

@@

In addition to the base text, semantic authoring also has to provide the semantic
annotations that drive the personalization and interaction downstream. Techniques from formal
authoring could help here, but in contrast to formal authoring, we cannot just trigger the
IDE functionality by the formal syntax, as the authored content is -- so far -- informal
and the content commons (the target of the annotations) is itself only flexiformal. In a
way, we need a local flexiformalization workflow that helps authors bridge the
informal-formal gap.

A distinctive feature of semantic authoring that is not apparent at the first glance is
that the added investment in annotation moves the sweet spot in the re-use/customize
trade-off towards sharing and interoperability. In our experience sufficiently annotated
documents can be automatically customized if the semantics has multiple
notations/verbalizations/presentations. Consider for instance course materials on complex
arithmetic, which become re-usable across the electrical engineering/mathematics border if
the imaginary unit in formulae is represented e.g. by a semantic macro
\lstinline|\imaginaryunit|, which can alternatively \edited{be presented as} $i$ (for uses in maths) or
$j$ (for EE uses). As a consequence semantic authoring is more geared towards creating
reusable/shareable document fragments organized in central or federated repositories and
hubs. This is quite similar to the formal authoring arena, where we see a tendency towards
massively shared math libraries in formal systems and package/library hubs for programming
libraries. Semantic authoring support systems should therefore facilitate the
logistics of using and curating shared libraries.

@@

How the annotation process can be supported depends on the particular semantic framework,
the specific annotations we want to support, and the semantics embedded in the content
commons.  In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers
to their definienda.  A tool for supporting semantic authoring of formulae via notations
has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24},
\edited{so we disregard this important aspect here.
The tool} uses a grammar to parse informal, presentation-focused formulae (e.g.\
\lstinline[keywordstyle={}]|A\times B|) into abstract syntax trees based on semantic
macros defined in \sTeX, which can then replace the original formula (e.g.\
\lstinline|\cart{A,B}| for the Cartesian product of sets).  In the case of ambiguity,
authors can pick the correct reading in a graphical user interface that visualizes the
different abstract syntax trees.

@@

To skip the word, the user can type \lstinline|s|;
\lstinline|s!| skips it until the end of the file, and \lstinline|S|
will skip it in future runs as well by appending a \snify-specific comment to the file.
There are numerous other commands, e.g.\ to adjust the selection,
search for alternative annotation targets,
or view/edit the file introducing one of the choices.

@@

\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents.}
As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
documents that were originally authored without semantic annotations.  It can be
configured to consider different \LaTeX\ macros to identify definienda (e.g.\
\lstinline[language={}]|\emph| or \lstinline[language={}]|\textbf|), and then steps
through them as \snify does.
For each potential definiendum, the user can then use fuzzy
search to find the corresponding symbol in the domain model
\edited{and replace the original macro with an} \sTeX \edited{definiendum annotation}.
This populates the catalog
with verbalizations relevant to the document, making subsequent \snify runs more
effective, especially for documents in languages that are not well covered by the domain
model (i.e.\ not English).

\paragraph{Evaluation}
% \section{Practical Evaluation and Future Work}\label{sec:eval}
\begin{wrapfigure}{r}{0.4\textwidth}
    \vskip-1.2cm
  \centering
  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
    \vskip-0.2cm
  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
  \vskip-0.7cm
\end{wrapfigure}
In our experience, \snify's step-through workflow for annotating term references 
% for symbols from the domain model 
is almost an order of magnitude more efficient than writing the
annotations and imports by hand.\footnote{
        The precise productivity gain depends on the annotator's
        familiarity with the domain model (\edited{searching and identifying}
        the right symbol is time-consuming) but also on other factors like
        the typing speed.
        Measurements are complicated by the fact that
        authors tend to annotate more terms when using \snify.
        We have observed speed-ups over IDE-based annotation
        ranging from 3 (for an expert knowing the domain model by heart)
        to 13 (for a beginner).
        The focus mode further increases productivity
        when considering the overall corpus quality.
}
% , even for annotators who are familiar with the domain model.
Given the sheer number of term annotations (see \Cref{fig:annocounts}),
this results in a considerable productivity gain.
For annotators unfamiliar with the domain model, the unassisted annotation task is
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices
% \ednote{
%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
% }
-- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

For less complete domain models we have to skip too many terms that should ultimately be
annotated and annotation effectiveness suffers. This is currently the case for all
non-English languages in the \sTeX corpus we work with. Coverage in German is only about
half of that of English in the domain model and we can already see the practical effects.  The \defianno tool
can mitigate this partially by providing verbalizations for
the definienda in the document, but it does not help with terms that are not explicitly
defined in the document.

@@

\snify's interface is inspired by traditional spell checkers
like \lstinline|ispell|~\cite{ispellman}
that step through documents and suggest corrections for misspelled words,
\edited{which is of course a very different underlying task.}

The bulk annotation workflow for existing informal text is also relevant when creating
machine learning datasets, e.g.\ for named entity recognition.  There are many (often
browser-based) tools for this task; a prominent example is
\textsf{brat}~\cite{brat:on} for text annotation.
\edited{
    Technical documents with tables, complex formulae, etc. cannot
    easily be expressed as sequences of words and require different annotation
    tools like} e.g.\ \textsf{AnnoTize}~\cite{PanzerSchaefer:AnnoTize23}
\edited{
    that supports the annotation of HTML documents containing MathML formulae.
}

In all cases, the setting, solution, and resulting workflows differ from the one presented
in this paper. The WissKI system \cite{goerz2010adaptation} has a rudimentary text edition
facility that allows annotating documents with ontology -- mostly ABox -- references; this
may be the nearest match to our \snify system.

\section{Conclusion}\label{sec:conclusion}
In this paper we have presented the semantic authoring task in a flexiformal context
using the \sTeX corpus and the adaptive learning assistant \ALeA as a concrete example. We
have shown that it differs from both classical authoring and formal authoring in terms of
the necessary system support. As an example of specialized authoring support, we have
presented \snify, a command-line tool that bridges the gap between informal text and
formal/semantic symbols in \sTeX.
\edited{It is open source; the code and documentation are available at}
\cite{stextools:git}.
Given the impact of \snify on the authoring process, we are
currently working on a deeper integration of \snify into the \sTeX ecosystem and content
commons as described in \Cref{sec:scaling}.
